---
  title: "UCell Analysis Tutorial on Cricket Spatial Data"
author: "Bioinformatics Lab"
date: "`r Sys.Date()`"
output: html_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Load Required Libraries

```{r libraries}
library(UCell)
library(Matrix)
library(Seurat)
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(jsonlite)
library(png)
```

## 2. Set Paths and Load Seurat Object

```{r paths}
base_path <- "../../datashare/grillus"
data_path <- file.path(base_path, "files")
output_dir <- file.path("..", "results", "ucell")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

seurat_obj <- readRDS(file.path(base_path, "grillus_seurat.rds"))
```

## 3. Load Gene Signatures from CSV Files

```{r gene-signatures-csv}
read_gene_list <- function(file, gene_col = 1) {
  df <- read_csv(file)
  genes <- df[[gene_col]]
  tolower(str_trim(gsub("_", "-", as.character(genes))))
}

genes_growth     <- read_gene_list(file.path(data_path, "de_gene_names_growth.csv"), gene_col = "gene_id")
genes_regen_up   <- read_gene_list(file.path(data_path, "ur_regeneration.csv"), gene_col = "gene_name")
genes_regen_down <- read_gene_list(file.path(data_path, "dr_regeneration.csv"), gene_col = "gene_name")
genes_scarring   <- read_gene_list(file.path(data_path, "gene_name_scaring.csv"), gene_col = "gene_id")

# Annotate gene features
gene_annot <- seurat_obj@misc$gene_annotations

add_flag <- function(genes) rownames(gene_annot) %in% genes

gene_annot$growth_gene     <- add_flag(genes_growth)
gene_annot$regen_up_gene   <- add_flag(genes_regen_up)
gene_annot$regen_down_gene <- add_flag(genes_regen_down)
gene_annot$scarring_gene   <- add_flag(genes_scarring)

seurat_obj@misc$gene_annotations <- gene_annot
```

## 4. Calculate UCell Scores for CSV Gene Sets

```{r ucell-csv}
signatures <- list(
  growth     = rownames(gene_annot)[gene_annot$growth_gene],
  regen_up   = rownames(gene_annot)[gene_annot$regen_up_gene],
  regen_down = rownames(gene_annot)[gene_annot$regen_down_gene],
  scarring   = rownames(gene_annot)[gene_annot$scarring_gene]
)

for (sig in names(signatures)) {
  seurat_obj <- AddModuleScore_UCell(
    seurat_obj,
    features = list(signatures[[sig]]),
    name = paste0(sig, "UCell")
  )
}
```

## 5. Load and Attach Spatial Image

```{r image}
spatial_path <- file.path(base_path, "outs", "spatial")
img_obj <- Read10X_Image(image.dir = spatial_path, filter.matrix = TRUE)
colnames(img_obj) <- colnames(seurat_obj)

DefaultImage(seurat_obj) <- "image"
seurat_obj[["image"]] <- img_obj
DefaultAssay(seurat_obj) <- "Spatial"
```

## 6. Visualize UCell Scores

```{r plot-csv-scores}
features_to_plot <- c("growthUCell", "regen_upUCell", "regen_downUCell", "scarringUCell")

for (feat in features_to_plot) {
  p <- SpatialFeaturePlot(
    seurat_obj,
    features = feat,
    pt.size.factor = 3
  ) + ggtitle(feat)
  
  ggsave(
    filename = file.path(output_dir, paste0(feat, ".png")),
    plot = p,
    width = 6,
    height = 6,
    dpi = 300
  )
}
```

## 7. Additional Gene Sets from PDF (Pathways & Regulators)

```{r pdf-signatures}
# Example manually annotated gene sets from database list (mapped to gene IDs beforehand)
signatures_interest <- list(
  EGFR = c("gbi-12345", "gbi-12346"),
  Dpp_TGFb = c("gbi-23456", "gbi-23457"),
  Insulin_IGF = c("gbi-34567", "gbi-34568"),
  Notch = c("gbi-45678", "gbi-45679"),
  Toll = c("gbi-56789", "gbi-56780"),
  TFs = c("gbi-67890", "gbi-67891"),
  Methyl_Histone = c("gbi-78901", "gbi-78902"),
  Chromatin_Remodeling = c("gbi-89012", "gbi-89013")
)

for (sig in names(signatures_interest)) {
  seurat_obj <- AddModuleScore_UCell(
    seurat_obj,
    features = list(signatures_interest[[sig]]),
    name = paste0("signature_1", sig)
  )
}
```

## 8. Visualize database-Informed Scores

```{r plot-flybase-scores}
ucell_feats <- grep("^signature_1", colnames(seurat_obj@meta.data), value = TRUE)

for (feat in ucell_feats) {
  p <- SpatialFeaturePlot(
    seurat_obj,
    features = feat,
    pt.size.factor = 3
  ) + ggtitle(feat)
  
  ggsave(
    filename = file.path(output_dir, paste0("binary/", feat, ".png")),
    plot = p,
    width = 6,
    height = 6,
    dpi = 300
  )
}
```

## 10. Visualize database-Informed Scores (Binary)

```{r plot-flybase-scores-binary}
for (feat in ucell_feats) {
  bin_feat <- seurat_obj@meta.data[[feat]] > 0
  seurat_obj@meta.data[[paste0(feat, "_bin")]] <- factor(as.integer(bin_feat))

  p <- SpatialFeaturePlot(
    seurat_obj,
    features = paste0(feat, "_bin"),
    pt.size.factor = 3
  ) + ggtitle(paste0(feat, " (Binary)"))

  ggsave(
    filename = file.path(binary_dir, paste0(feat, "_binary.png")),
    plot = p,
    width = 6,
    height = 6,
    dpi = 300
  )
}
```

## 11. Save Final Seurat Object

```{r save-final}
saveRDS(seurat_obj, file.path(base_path, "grillus_seurat_ucell_final.rds"))
```

---

**Note:** Gene IDs are placeholders; replace with real mapped values when running on your dataset.

**Next Steps:**
- Add biological interpretation layers
- Cross-species ontology mapping (e.g., via FlyBase, GO)
- Combine UCell with clustering, pseudotime or region segmentation

